### 2024-09-01 cognitive-bias.qmd
20240901 - law of small numbers

The gambler's fallacy occurs due to an erroneous belief in the law of small numbers.
The law of large numbers is a mathematical theorem that the average of a sufficiently large number of independent observations converges to the true value.
For instance, if you flip a fair coin 1 million times, it is likely to land heads-up ~50% of the time.
The law of *small* numbers (aka hasty generalization), by contrast, is an erroneous belief that small samples are representative of the populations from which they were drawn.
For instance, if you flip a coin 10 times, belief in the law of small numbers would lead one to believe that the coin will flip heads-up exactly 5 times out of 10.
However, in reality, the chance is less than 1 in 4 (24.6%) that exactly 5 of 10 coin flips turn up heads, as calculated below and as depicted in @fig-simulationOf10CoinFlips:

@@

numHeads <- rbinom(
  n = 100000,
  size = 10,
  prob = .5
)

simulationOfFlipping10Coins <- data.frame(
  numHeads = numHeads
)

simulationOfFlipping10Coins <- simulationOfFlipping10Coins %>% 
  mutate(
    highlight = ifelse(numHeads == 5, "yes", "no")
  )

ggplot2::ggplot(
  data = simulationOfFlipping10Coins,
  mapping = aes(
    x = numHeads,
    fill = highlight)
) +
  geom_histogram(
    color = "#000000",
    bins = 11
  ) +
  scale_x_continuous(
    breaks = 0:10
  ) +
  scale_fill_manual(
    values = c(
      "yes" = "tomato",
      "no" = "gray")
  ) +
  labs(
    x = "Number of Coins Flipped Heads (out of 10 Coin Flips)",
    y = "Frequency",
    title = "Histogram of Number of Coins that Flip Heads-Up\nin a Simulation of 10 Coin Flips\n(with 100,000 Replications)."
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

@@

>>>>

The gambler's fallacy occurs due to an erroneous belief in the law of small numbers.
The law of large numbers is a mathematical theorem that the average of a sufficiently large number of independent observations converges to the true value.
For instance, if you flip a fair coin 1 million times, it is likely to land heads-up ~50% of the time.
The law of *small* numbers (aka hasty generalization), by contrast, is an erroneous belief that small samples are representative of the populations from which they were drawn.
In general, large samples yield more precise results than small samples, whereas extreme results (in either direction—e.g., much higher or much lower than 50% of coins that flip heads-up) are more likely to be obtained from small samples [@Kahneman2011].
For instance, if you flip a coin 10 times, belief in the law of small numbers would lead one to believe that the coin will flip heads-up exactly 5 times out of 10.
However, in reality, the chance is less than 1 in 4 (24.6%) that exactly 5 of 10 coin flips turn up heads, as calculated below and as depicted in @fig-simulationOf10CoinFlips:

@@

smallNumFlips <- 10
largeNumFlips <- 1000000
numSimulations <- 100000

numHeadsSmallNumFlips <- rbinom(
  n = numSimulations,
  size = smallNumFlips,
  prob = .5
)

numHeadsLargeNumFlips <- rbinom(
  n = numSimulations,
  size = largeNumFlips,
  prob = .5
)

simulationOfFlippingCoins <- data.frame(
  numHeadsSmallNumFlips = numHeadsSmallNumFlips,
  numHeadsLargeNumFlips = numHeadsLargeNumFlips
)

simulationOfFlippingCoins <- simulationOfFlippingCoins %>% 
  mutate(
    proportionHeadsSmallNumFlips = numHeadsSmallNumFlips / smallNumFlips,
    proportionHeadsLargeNumFlips = numHeadsLargeNumFlips / largeNumFlips,
    highlight = ifelse(numHeads == 5, "yes", "no"),
    extremeSmallNumFlips = ifelse(proportionHeadsSmallNumFlips <= 0.2 | proportionHeadsSmallNumFlips >= 0.8, TRUE, FALSE),
    extremeLargeNumFlips = ifelse(proportionHeadsLargeNumFlips <= 0.2 | proportionHeadsLargeNumFlips >= 0.8, TRUE, FALSE),
  )

ggplot2::ggplot(
  data = simulationOfFlippingCoins,
  mapping = aes(
    x = numHeadsSmallNumFlips,
    fill = highlight)
) +
  geom_histogram(
    color = "#000000",
    bins = 11
  ) +
  scale_x_continuous(
    breaks = 0:10
  ) +
  scale_fill_manual(
    values = c(
      "yes" = "tomato",
      "no" = "gray")
  ) +
  labs(
    x = "Number of Coins Flipped Heads (out of 10 Coin Flips)",
    y = "Frequency",
    title = "Histogram of Number of Coins that Flip Heads-Up\nin a Simulation of 10 Coin Flips\n(with 100,000 Replications)."
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

@@

Moreover, among the simulations with only 10 coin flips, nearly 11% of sets had extreme results (≤ 20% heads or ≥ 80% heads):

```{r}
prop.table(table(simulationOfFlippingCoins$extremeSmallNumFlips))
```

By contrast, among the simulations with 1 million coin flips, none of the sets had extreme results:

```{r}
prop.table(table(simulationOfFlippingCoins$extremeLargeNumFlips))
```
